\chapter{Trigonometry Review}
\label{chap:TR}
Further reading: \texttt{Chen \& Duong ``mtfym01.pdf''.}

\begin{bigideas}{sec:TR Big Ideas}
\begin{itemize}
  \item Radians and arc length are interrelated. There are $2\pi$ radians
  in a circle.
  \item An angle subtended by $\theta$ will have arc length $\theta$.
  \item $\cos(\theta) = x$-ordinate, $\sin(\theta) = y$-ordinate
  \item $\tan(\theta) = \frac{\sin(\theta)}{\cos(\theta)} = \frac{y}{x}$
  \item \nameref{sec:TR Pythagorean Identities} (page \pageref{sec:TR
  Pythagorean Identities}) are worth learning!
\end{itemize}
\end{bigideas}

\section{Radian \& Arc Length}
\label{sec:TR Radian and Arc Length}
Further reading: \texttt{Chen \& Duong ``mtfym01.pdf'' pp 1-2.}

\section{Trigonometric Functions}
\label{sec:TR Trigonometric Functions}
Further reading: \texttt{Chen \& Duong ``mtfym01.pdf'' pp 2-4.}

\section{Pythagorean Identities}
\label{sec:TR Pythagorean Identities}
Further reading: \texttt{Chen \& Duong ``mtfym01.pdf'' pp 4-5.}

\subsection{Properties of Trigonometric Functions}
\label{subsec:TR Trigonometric Functions - Properties}
Further reading: \texttt{Chen \& Duong ``mtfym01.pdf'' pp 5-10.}

\section{Trigonometric Identities}
\label{sec:TR Trigonometric Identities}
Further reading: \texttt{Chen \& Duong ``mtfym01.pdf'' pp 11.}

\subsection{Sine Rule}
\label{subsec:TR Trigonometric Identities - Sine Rule}

\subsection{Cosine Rule}
\label{subsec:TR Trigonometric Identities - Cosine Rule}

\subsection{Angle Sum \& Difference Identities}
\label{subsec:TR Trigonometric Identities - Angle Sum Identities}
Further reading: \texttt{Chen \& Duong ``mtfym01.pdf'' pp 12.}

\subsubsection{Double Angle Identity}
\label{subsubsec:TR Trigonometric Identities - Angle Sum Identities - Double Angle Identity}
Further reading: \texttt{Chen \& Duong ``mtfym01.pdf'' pp 12.}

\subsubsection{Half Angle Identity}
\label{subsubsec:TR Trigonometric Identities - Angle Sum Identities - Half Angle Identity}
Further reading: \texttt{Chen \& Duong ``mtfym01.pdf'' pp 13-15.}